---
title: "Practical Machine Learning: Prediction Assignment Writeup"
author: "Karan Shishoo"
date: "April 28, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).

## Data

The training data for this project are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

The data for this project come from this source: <http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har>. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## Process

### Reproducibility

A random seed (42819) was selected for the code to ensure reproducibility In order to validate the results produced below use the same seed.

### Model building

The outcome variable ("Classe") has five types, as noted from source^[<http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har>] -
Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions:

* exactly according to the specification (Class A)
* throwing the elbows to the front (Class B)
* lifting the dumbbell only halfway (Class C)
* lowering the dumbbell only halfway (Class D)
* throwing the hips to the front (Class E)

The predictions  will be evaluated by maximizing the accuracy  and minimizing  the out of sample error. All possible variables available  after cleaning will be used to build the model. Both decision tree and random forest algorithms will be used as they excel at identifying important features for classification. The better model of the two will be selected.

### Cross-Validation

Cross-Validation will be performed to ensure that our model is not overfit and so that the most accurate model will be selected The training data will be split 7:3 to create a subTraining set and a subTesting set for this.

## Initial Setup Code

Installing libraries and setting seed
```{r warning=FALSE, error=FALSE}
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(corrplot)
library(RColorBrewer)
library(rattle)
set.seed(42819)
```

Downloading the data

```{r warning=FALSE, error=FALSE}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"

if (!file.exists("./data")) {
  dir.create("./data")
}

if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile = trainFile, method = "curl")
}

if (!file.exists(testFile)) {
  download.file(testUrl, destfile = testFile, method = "curl")
}

training <- read.csv(trainFile)
testing <- read.csv(testFile)

rm(trainFile)
rm(testFile)
```


Cleaning the Data (Removing ID)
```{r warning=FALSE, error=FALSE}
training <- training[c(-1)]
```

Cleaning the Data (Near Zero Variance)

```{r warning=FALSE, error=FALSE}
DataNZV <- nearZeroVar(training, saveMetrics=TRUE)
training <- training[, !DataNZV$nzv]
rm(DataNZV)
```

Cleaning the Data (Contains NA's or not much contribution)

```{r warning=FALSE, error=FALSE}
condition <- (colSums(is.na(training)) == 0)
training <- training[, condition]

regex <- grepl("^X|timestamp|user_name", names(training))
training <- training[, !regex]

rm(condition)
rm(regex)
```

Correlation Matrix:
```{r warning=FALSE, error=FALSE}
corrplot(cor(training[, -length(names(training))]), method = "color", tl.cex = 0.5)
```


Splitting the training data

```{r warning=FALSE, error=FALSE}
inTrain <- createDataPartition(y=training$classe, p=0.70, list=FALSE)
subTraining <- training[inTrain, ]; subTesting <- training[-inTrain, ]
rm(inTrain)
```

## Data Modeling


### Decision Tree model
```{r warning=FALSE, error=FALSE}
modelDTree <- rpart(classe ~ ., data = subTraining, method = "class")
fancyRpartPlot(modelDTree)
```

### Decision Tree validation

```{r warning=FALSE, error=FALSE}
predictDTree <- predict(modelDTree, subTesting, type = "class")
confusionMatrix(subTesting$classe, predictDTree)
accuracy <- postResample(predictDTree, subTesting$classe)
ose <- 1 - as.numeric(confusionMatrix(subTesting$classe, predictDTree)$overall[1])
rm(predictDTree)
rm(modelDTree)
```  
The Estimated accuracy  is `r accuracy[1]*100`% while the out of sample error is `r ose*100`%. 

### Random Forest model
[done using 5-fold cross validation]
```{r warning=FALSE, error=FALSE}
modelRF <- train(classe ~ ., data = subTraining, method = "rf", trControl = trainControl(method = "cv", 5), ntree = 250)
modelRF
```  

### Random Forest validation

```{r warning=FALSE, error=FALSE}
predictRF <- predict(modelRF, subTesting)
confusionMatrix(subTesting$classe, predictRF)
accuracy <- postResample(predictRF, subTesting$classe)
ose <- 1 - as.numeric(confusionMatrix(subTesting$classe, predictRF)$overall[1])
rm(predictRF)
```
The Estimated accuracy   is `r accuracy[1]*100`% while the out of sample error is `r ose*100`%. 

##Test set prediction and output
From the estimated accuracies  it is evident that Random Forest is the more optimal model, thus it is used in generating the predictions for the test set.

```{r warning=FALSE, error=FALSE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}

pml_write_files(predict(modelRF, testing[, -length(names(testing))]))

rm(accuracy)
rm(ose)
rm(modelRF)
rm(training)
rm(testing)
rm(validation)
rm(pml_write_files)
```  

